---
title: 'Multivariate analysis: multiple logistic regression'
author: Núria Jolis Orriols
date: '`r format(Sys.Date(),"%e de %B, %Y")`' 
output:
  pdf_document:
    toc: yes
    df_print: kable
header-includes:
  - \usepackage[english]{babel}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NULL, cache=TRUE, error = TRUE)
options(width=90)
```

```{r packages, message=FALSE, echo=FALSE, warning=FALSE}
libraries <- c("kernlab", "ggplot2" ,"caret", "tidyverse", "dplyr", "gmodels", "MASS")
check.libraries <- is.element(libraries, installed.packages()[, 1])==FALSE
libraries.to.install <- libraries[check.libraries]
if (length(libraries.to.install!=0)) {
  install.packages(libraries.to.install)
}

success <- sapply(libraries,require, quietly = FALSE,  character.only = TRUE)
if(length(success) != length(libraries)) {stop("A package failed to return a success in require() function.")}
```

# Step 1 - Obtaining datasets

```{r}
dataA <- read.csv("data/datasetA.csv",stringsAsFactors=T)
dataB <- read.csv("data/datasetB.csv",stringsAsFactors=T)
dataC <- read.csv("data/datasetC.csv",stringsAsFactors=T)
dataD <- read.csv("data/datasetD.csv",stringsAsFactors=T)
dataE <- read.csv("data/datasetE.csv",stringsAsFactors=T)
```

# Step 2 - Generate multiple logistic regresion models 

## Data A

### Full logistic regression model

```{r}
xnames <- names(dataA[2:48])
foriA<- as.formula(paste("outcome ~ ", paste(xnames, collapse= "+")))
```


```{r}
full_mlrA<- glm(foriA, data = dataA, family = 'binomial')
summary(full_mlrA)
```
The full model with dataset A found the variables age, presence of deficiencyanemia, presence of renal.failure, temperature, platelets and PCO2 to be statistically significant.

### Perform stepwise variable selection

The stepAIC function of the MASS package will be used to select the most contributive variables:

```{r, message=FALSE, warning=FALSE}
step_mlrA<- full_mlrA %>% stepAIC(trace = FALSE)
```
The model with the stepAIC function does not converge because of the missing values. 

## Data B

### Full logistic regression model

```{r}
xnamesB <- names(dataB[2:48])
foriB<- as.formula(paste("outcome ~ ", paste(xnamesB, collapse= "+")))
```


```{r}
full_mlrB<- glm(foriB, data = dataB, family = 'binomial')
summary(full_mlrB)
```

The full model found the variables age, deficiencyanemia, renal.failure, temperature, platelets and PCO2 to be statistically significant.

### Perform stepwise variable selection

The stepAIC function of the MASS package will be used to select the most contributive variables:

```{r, message=FALSE, warning=FALSE}
step_mlrB <- full_mlrB %>% stepAIC(trace = FALSE)
summary(step_mlrB)
```

The function chose a final model with the following formula: 

outcome ~ age + BMI + atrialfibrillation + deficiencyanemias + 
    renal.failure + COPD + heart.rate + temperature + leucocyte + 
    platelets + neutrophils + lymphocyte + PT + urea.nitrogen + 
    glucose + blood.calcium + chloride + anion.gap + magnesium.ion + 
    PCO2
 
It found the variables age, deficiencyanemia, renal.failure, temperature, SP.O2, platelets, leucocyte, neutrophils, lymphocyte, blood.calcium, blood.potassium, bicarbonate, pH and PCO2 to be statistically significant.

```{r}
formula(step_mlrB)
results_mlrB<-cbind( "OR" = exp(coef(step_mlrB)), "p-value" = coef(summary(step_mlrB))[,4])
results_mlrB
```


## Data C

### Full logistic regression model

```{r}
xnamesC <- names(dataC[2:40])
foriC<- as.formula(paste("outcome ~ ", paste(xnamesC, collapse= "+")))
```


```{r}
full_mlrC<- glm(foriC, data = dataC, family = 'binomial')
summary(full_mlrC)
```

The full model found the variables age, deficiencyanemia, renal.failure, platelets, leucocyte, blood.potassium, blood.calcium and magnesium.ion to be statistically significant.

### Perform stepwise variable selection

The stepAIC function of the MASS package will be used to select the most contributive variables:

```{r, message=FALSE,warning=FALSE}
step_mlrC <- full_mlrC %>% stepAIC(trace = FALSE)
```
The model with the stepAIC function does not converge because of the missing values. 

## Data D

### Full logistic regression model

```{r, message=FALSE}
xnamesD <- names(dataD[2:48])
foriD<- as.formula(paste("outcome ~ ", paste(xnamesD, collapse= "+")))
```


```{r}
full_mlrD<- glm(foriD, data = dataD, family = 'binomial')
summary(full_mlrD)
```

The full model found the variables age, deficiencyanemia, renal.failure, temperature, platelets and PCO2 to be statistically significant.

### Perform stepwise variable selection

The stepAIC function of the MASS package will be used to select the most contributive variables:

```{r, message=FALSE,warning=FALSE}
step_mlrD <- full_mlrD %>% stepAIC(trace = FALSE)
summary(step_mlrD)
```

The function chose a final model with the following formula: 

outcome ~ age + BMI + deficiencyanemias + renal.failure + 
    COPD + heart.rate + temperature + SP.O2 + RDW + leucocyte + 
    platelets + lymphocyte + urea.nitrogen + blood.potassium + 
    blood.sodium + blood.calcium + chloride + magnesium.ion + 
    bicarbonate + lactic.acid + PCO2 + m.a.p
 
It found the variables age, deficiencyanemia, renal.failure, temperature, SP.O2, platelets, leucocyte, neutrophils, lymphocyte, blood.calcium, blood.potassium, bicarbonate, pH and PCO2 to be statistically significant.

```{r}
formula(step_mlrD)
results_mlrD<-cbind( "OR" = exp(coef(step_mlrD)), "p-value" = coef(summary(step_mlrD))[,4])
results_mlrD
```


## Data E

### Full logistic regression model

```{r}
xnamesE <- names(dataE[2:48])
foriE<- as.formula(paste("outcome ~ ", paste(xnamesE, collapse= "+")))
```


```{r}
full_mlrE<- glm(foriE, data = dataE, family = 'binomial')
summary(full_mlrE)
```

The full model found the variables age, deficiencyanemia, renal.failure, temperature, platelets and PCO2 to be statistically significant.

### Perform stepwise variable selection

The stepAIC function of the MASS package will be used to select the most contributive variables:

```{r, message=FALSE,warning=FALSE}
step_mlrE <- full_mlrE %>% stepAIC(trace = FALSE)
summary(step_mlrE)
```

The function chose a final model with the following formula: 

outcome ~ age + BMI + atrialfibrillation + deficiencyanemias + 
    renal.failure + COPD + heart.rate + temperature + leucocyte + 
    platelets + neutrophils + lymphocyte + PT + urea.nitrogen + 
    glucose + blood.calcium + chloride + anion.gap + magnesium.ion + 
    PCO2
 
It found the variables age, deficiencyanemia, renal.failure, temperature, SP.O2, platelets, leucocyte, neutrophils, lymphocyte, blood.calcium, blood.potassium, bicarbonate, pH and PCO2 to be statistically significant.

```{r}
formula(step_mlrE)
results_mlrE<-cbind( "OR" = exp(coef(step_mlrE)), "p-value" = coef(summary(step_mlrE))[,4])
results_mlrE
```
