---
title: 'Comparison of RF models for outcome prediction of ICU-admitted heart failure patients'
author: Núria Jolis Orriols
date: '`r format(Sys.Date(),"%e de %B, %Y")`' 
output:
  pdf_document:
    toc: yes
    df_print: kable
header-includes:
  - \usepackage[english]{babel}
params:
  p.train: !r 0.7
  seed.train: 12345
  seed.clsfier: 1234567
  seed.improv: 123
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NULL, cache=TRUE, error = TRUE)
options(width=90)
```

```{r packages, message=FALSE, echo=FALSE, warning=FALSE}
libraries <- c("randomForest", "ggplot2" ,"caret", "tidyverse", "dplyr", "gmodels", "ROSE", "mice", "VIM")
check.libraries <- is.element(libraries, installed.packages()[, 1])==FALSE
libraries.to.install <- libraries[check.libraries]
if (length(libraries.to.install!=0)) {
  install.packages(libraries.to.install)
}

success <- sapply(libraries,require, quietly = FALSE,  character.only = TRUE)
if(length(success) != length(libraries)) {stop("A package failed to return a success in require() function.")}
```

# Step 1 - Obtaining data

```{r}
dataA <- read.csv("data/datasetA.csv",stringsAsFactors=T)
dataB <- read.csv("data/datasetB.csv",stringsAsFactors=T)
dataC <- read.csv("data/datasetC.csv",stringsAsFactors=T)
dataD <- read.csv("data/datasetD.csv",stringsAsFactors=T)
dataE <- read.csv("data/datasetE.csv",stringsAsFactors=T)
```

# Step 2 - Model training

## Data training/test partition into 70-30%

```{r}
#Data A
set.seed(12345)
nA<-nrow(dataA)
trainA<-sample(nA,floor(nA*params$p.train))
dataA_train<-dataA[trainA,]
dataA_test<-dataA[-trainA,]
```

```{r}
#Data B
set.seed(12345)
nB<-nrow(dataB)
trainB<-sample(nB,floor(nB*params$p.train))
dataB_train<-dataB[trainB,]
dataB_test<-dataB[-trainB,]
```

```{r}
#Data C
nC<-nrow(dataC)
trainC<-sample(nC,floor(nC*params$p.train))
dataC_train<-dataC[trainC,]
dataC_test<-dataC[-trainC,]
```

```{r}
#Data D
set.seed(12345)
nD<-nrow(dataD)
trainD<-sample(nD,floor(nD*params$p.train))
dataD_train<-dataD[trainD,]
dataD_test<-dataD[-trainD,]
```

```{r}
#Data E
set.seed(12345)
nE<-nrow(dataE)
trainE<-sample(nE,floor(nE*params$p.train))
dataE_train<-dataE[trainE,]
dataE_test<-dataE[-trainE,]
```

## Model training

We are going to generate RF models with the training data to predict the outcome. We use the randomForest() function of the randomForest package. 

### Data A

```{r,warning=FALSE,message=FALSE}
rf.A<-randomForest(outcome ~ ., data=dataA_train)
rf.A
```

### Model B

```{r,warning=FALSE,message=FALSE}
rf.B<-randomForest(outcome ~ ., data=dataB_train)
rf.B
```

```{r}
plot(rf.B)
importance(rf.B)
varImpPlot(rf.B)
```

### Model C

```{r,warning=FALSE,message=FALSE}
rf.C<-randomForest(outcome ~ ., data=dataC_train)
rf.C
```
### Model D

```{r,warning=FALSE,message=FALSE}
rf.D<-randomForest(outcome ~ ., data=dataD_train)
rf.D
```

```{r}
plot(rf.D)
importance(rf.D)
varImpPlot(rf.D)
```

### Model E

```{r,warning=FALSE,message=FALSE}
rf.E<-randomForest(outcome ~ ., data=dataE_train)
rf.E
```

```{r}
plot(rf.E)
importance(rf.E)
varImpPlot(rf.E)
```

## Step 4 - Model prediction and evaluation

**Prediction**

We use the function `predict` to generate predictions on the testing dataset:


### B
```{r}
pred.rf.B<-predict(rf.B, dataB_test[-1])
```

### D
```{r}
pred.rf.D<-predict(rf.D, dataD_test[-1])
```

### E
```{r}
pred.rf.E<-predict(rf.E, dataE_test[-1])
```

**Evaluation**

### B
```{r}
confusionMatrix(dataB_test$outcome, pred.rf.B)
roc.curve(dataB_test$outcome, pred.rf.B)
```
### D
```{r}
confusionMatrix(dataD_test$outcome,pred.rf.D)
roc.curve(dataD_test$outcome,pred.rf.D)
```

### E
```{r}
confusionMatrix(dataE_test$outcome,pred.rf.E)
roc.curve(dataE_test$outcome,pred.rf.E)
```

## Step 5 - Improving model performance

To improve model's performance a k-fold cross-validation method will be used. It consist in a resampling method that uses different portions of the data to test and train a model on different iterations. 

```{r}
set.seed(params$seed.improv)
ctrl <- trainControl(method='repeatedcv',
                            number=10,
                            summaryFunction = defaultSummary,
                            verboseIter = FALSE,
                            repeats = 3)
grid_rf<- expand.grid(.mtry=c(2,4,8,16))
```

### Dataset B

```{r}
model.rf.B.Kfold<- train(outcome ~ ., 
                          data=dataB_train,
                          method="rf",
                          trControl=ctrl,
                            tuneLength=9,
                          tuneGrid=grid_rf,
                          metric="Accuracy",
                          prePoc = c("center", "scale"),
                          verbose=FALSE,
                          trace=FALSE
                  )
```

```{r}
prdclassB<-predict(model.rf.B.Kfold, newdata=dataB_test)
str(prdclassB)
confusionMatrix(data=prdclassB, dataB_test$outcome)
roc.curve(dataB_test$outcome, prdclassB)
```

### Dataset D

```{r}
model.rf.D.Kfold<- train(outcome ~ ., 
                          data=dataD_train,
                          method="rf",
                          trControl=ctrl,
                            tuneLength=9,
                          tuneGrid=grid_rf,
                          metric="Accuracy",
                          prePoc = c("center", "scale"),
                          verbose=FALSE,
                          trace=FALSE
                  )
```

```{r}
prdclassD<-predict(model.rf.D.Kfold, newdata=dataD_test)
str(prdclassD)
confusionMatrix(data=prdclassD, dataD_test$outcome)
roc.curve(dataD_test$outcome, prdclassD)
```

### Dataset E

```{r}
model.rf.E.Kfold<- train(outcome ~ ., 
                          data=dataE_train,
                          method="rf",
                          trControl=ctrl,
                            tuneLength=9,
                          tuneGrid=grid_rf,
                          metric="Accuracy",
                          prePoc = c("center", "scale"),
                          verbose=FALSE,
                          trace=FALSE
                  )
```

```{r}
prdclassE<-predict(model.rf.E.Kfold, newdata=dataE_test)
str(prdclassE)
confusionMatrix(data=prdclassE, dataE_test$outcome)
roc.curve(dataE_test$outcome, prdclassE)
```
